Robustness of MinD oscillation in \emph{Escherichia coli} with diverse cell shapes

### 2014-08-05 paper/paper.tex
a few changes

\begin{figure*}
  \includegraphics[width=\textwidth]{../data/shape-p/plots/image-plot--p-300-50-0-0-1500.pdf}
  \caption{Countour plot images of the concentration of each protein
    species in a natural pill-shaped bacterium at regular intervals in
    time (one second intervals), with darker regions indicating higher
    concentration.  The cells pictured are $4\micron$ in length,
    measured from end to end.  The order of frames is such that
    individual MinD proteins begin at the bottom of the plot (in the
    MinD:ATP state in the cytoplasm), and progress upward until they
    reach the MinE:MinD:ATP membrane-bound complex.  At that point,
    they will spontaneously dissociate into cytoplasmic MinE (the top
    row) and the starting state of cytoplasmic MinD:ADP.  The
    densities plotted are integrated along the axis orthogonal to the
    page, and the color scale is chosen separately for each species
    with black as the maximum value over space and time.}
  \label{image-p}
\end{figure*}

We begin with the naturally occuring pill cell shape.  We peice this
shape together as two hemispherical endcaps attached on either end of
a cylinder.  This shape follows the early simulations of Huang
\emph{et al.} but differs in that we have added the end caps for a
more natural shape, expecting similar results.  We test cells of
radius $.5\micron$ and of lengths $2\micron$, $3\micron$, and
$5\micron$ measured from the tip of one end cap to the top of the
other. In each case, we observe quick establishement (within a few
periods) of extremely regular oscillatory behavoir that lasts
indefinitely.  The period increases with cell length, as shown in
Table~\ref{tab:pill-periods}, in agreement with the results of Huang
\emph{et al.}

\begin{table}
  \begin{tabular}{|r|c|c|c|l|}
    \hline
    Length($\mu$) & 2.50 & 3.00 & 4.00 & 5.00\\
    \hline
    Period(sec) & sim & 33 & 38 & 48 \\ \cline{2-2}
    \hline
  \end{tabular}
  \caption{Period of oscillation according to length of cell for
    pill-shaped cells.}\label{tab:pill-periods}\fixme{the 'sim' here
    means that it's waiting for a simulation.  Not sure where it went
    but I'm rerunning it.}
\end{table}

This simple cell shape is a good starting point for observing in
detail the dynamic interaction between the different stages of protein
that lead to their qualitatively distinct behavoir. Figure
\ref{image-p} shows the process in a series of frames taken from
simulation animations.  Each frame is 2.5 seconds ahead of the last,
and each image shows a the concentration of a given state of protein
summed over the coordinate orthogonal to the page.  The color scale
for each protein state is set according to the maximum density
observed.

@@

\section{Stability Analysis and Mannik's Experimental Shapes}
In general we have found that the protein reaction sequence discussed
above results in extremely robust oscillatory behavoir of polar
selection and oscillation in a wide variety of cell shapes.  We have
also found that while all sizes of cell will in time settle into
regular oscillations, there are size dependent distinctions in the
number and placement of the spatial positions of maximization.
Regular oscillatory behavoir between two poles at opposite ends of the
longer cellular direction is found in cells which are long enough in
one dimension for the reaction chain to be unstable but short enough
in the perpendicular dimension to be stable.  The limits are seen when
both dimensions are too short to be unstable, so that initial
assymetry in protein concentration relaxes into a motionless steady
state (see Figure~\ref{box-mannik}), or when both are too great, so
that the proteins exhibit more than two poles and those at extreme
postitions in both dimensions.

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{../data/shape-randst/plots/paper-arrow-plot}
  \caption{Arrow plots for our simulation and mannik's experimental}
  \label{arrow-compare-mannik}
\end{figure}

@@

\section{Triangular Shapes}

\begin{figure}
%  \includegraphics[width=4.2cm]{../data/shape-triangle/plots/box-plot_D--triangle-25-233-233-233-1500}
%  \includegraphics[width=4.2cm]{../data/shape-triangle/plots/box-plot_D--triangle-25-266-266-266-1500}
  \includegraphics[width=\columnwidth]{../data/shape-triangle/plots/box-plot_D--triangle-25-233-233-233-1500}
  \includegraphics[width=\columnwidth]{../data/shape-triangle/plots/box-plot_D--triangle-25-266-266-266-1500}
  %\includegraphics[width=4.2cm]{../data/shape-triangle/plots/box-plot_D--triangle-25-800-800-800-1500}
  \caption{Two box plots of triangle shapes.  The two smaller show
    sizes at the lower stability threshhold, one just below, one just
    above.  The larger triangle is the largest simulated. The vertical
    axis shows stacked and changing over time the total number of
    proteins that are of four different compound stages and in the
    different sections of the cell.}
  \label{box-triangle}
\end{figure}

Along with our other shapes we have chosen to simulate both
equilateral and iscoscolese triangular flattened shapes, since these
offer in a way the simplest shapes that are geometrically more
complicated than our pills.

Both triangle types show a similar oscillation pattern.  The
iscosolese exhibit regular oscillations characterized by one distinct
pole in the more accute angle of the three, and another 'pole' along
the opposite wall, peaking in the two more obtuse angles at the same
time.  The equilateral triangles settle very quickly into the same
pattern, where the choice of which is one pole corner is dependent
upon initial conditions.  As with the Mannik simulated shapes, there
is a critical small size below which the oscillations dampen out and
will not achieve a regular pattern.  Figure~\ref{box-triangle} shows
two plots of equilateral triangles.  The two smaller show the
largest that showed damped out behavoir and the smallest that
developed regular oscillations.  The larger shows the largest traingle
that we simulated, showing that at even very large sizes the behavoir
is consistent and stable.  Above the small size threshold, all
simulated triangles show regular patterns for as long as simulated
(roughly 800 seconds).

>>>>

\begin{figure*}
  \includegraphics[width=\textwidth]{../data/shape-p/plots/image-plot--p-300-50-0-0-1500.pdf}
  \includegraphics[width=\textwidth]{../data/shape-p/plots/single-image-stoch}
  \caption{Countour plot images of the concentration of each protein
    species in a natural pill-shaped bacterium at regular intervals in
    time (one second intervals), with darker regions indicating higher
    concentration. The upper plots shows results from the
    deterministic simulation and the lower shows results from the
    stochastic simulation.  The cells pictured are $4\micron$ in
    length, measured from end to end.  The order of frames is such
    that individual MinD proteins begin at the bottom of the plot (in
    the MinD:ATP state in the cytoplasm), and progress upward until
    they reach the MinE:MinD:ATP membrane-bound complex.  At that
    point, they will spontaneously dissociate into cytoplasmic MinE
    (the top row) and the starting state of cytoplasmic MinD:ADP.  The
    densities plotted are integrated along the axis orthogonal to the
    page, and the color scale is chosen separately for each species
    with black as the maximum value over space and time.  The
    stochastic densities are smeared out over space as guassians with
    characteristic length equal to the Abbe diffraction limit of
    $600nm$ wavelength light}
  \label{image-p}
\end{figure*}

We begin with the naturally occuring pill cell shape.  We peice this
shape together as two hemispherical endcaps attached on either end of
a cylinder.  This shape follows the early simulations of Huang
\emph{et al.} but differs in that we have added the end caps for a
more natural shape, expecting similar results.  We test cells of
radius $.5\micron$ and of lengths $2\micron$, $3\micron$, and
$5\micron$ measured from the tip of one end cap to the tip of the
other. In each case, we observe in the deterministic simulations a
quick establishement (within a few periods) of extremely regular
oscillatory behavoir that lasts indefinitely.  The stochastic results
show oscillitions that are similarly regular in time, with periods
that nearly match those of the deterministic (they differ roughly by a
factor of .05). The stochastic results also show a slight variation in
the height of the density maxima, but the peaks are still very well
defined. The period increases with cell length, as shown in
Table~\ref{tab:pill-periods}, in agreement with the results of Huang
\emph{et al.}.  The stochastic simulations show a similar pattern.
The MinD maxima occur at the ends of the cell in a back and forth
manner, and with the same approximate period.

\begin{table}
  \begin{tabular}{|r|c|c|c|l|}
    \hline
    Length($\mu$) & 2.50 & 3.00 & 4.00 & 5.00\\
    \hline
    Period(sec) & sim & 33 & 38 & 48 \\ \cline{2-2}
    \hline
  \end{tabular}
  \caption{Period of oscillation according to length of cell for
    pill-shaped cells.}\label{tab:pill-periods}
\end{table}

This simple cell shape is a good starting point for observing in
detail the dynamic interaction between the different stages of protein
that lead to their qualitatively distinct behavoir. Figure
\ref{image-p} shows the process in a series of frames taken from
simulation animations for both the deterministic and stochastic
simulations.  The stochastic concentrations have been 'smeared out' in
a manner meant to reproduce the images shown by a diffraction limited
microscope.  The protein densities which are located at specific grid
points give light to sorrounding grid points with a guassian weighting
function that has a variance equal to the Abbe diffraction limit of
$600nm$ wavelength light.  Each frame is 2.5 seconds ahead of the
last, and each image shows a the concentration of a given state of
protein summed over the coordinate orthogonal to the page.  The color
scale for each protein state is set according to the maximum values
observed.

@@

\section{Stability Analysis}
We have found that in both methods of simulation there is a lower
length scale that characterizes much of the behavoir.  Regular
oscillatory behavoir between two poles at opposite ends of the longer
cellular direction is found in cells which are long enough in one
dimension for the reaction chain to be unstable but short enough in
the perpendicular dimension to be stable.  The limits are seen when
both dimensions are too short to be unstable, so that initial
assymetry in protein concentration relaxes into a state of no
oscillation.  In these cell sizes, the deterministic simulations show
a motionless steady state and the stochastic simulations show random
fluctuations without any bulk behavoir (see Figure~\ref{box-mannik}).

@@

%% \begin{figure}
%%   \includegraphics[width=\columnwidth]{../data/shape-randst/plots/paper-arrow-plot}
%%   \caption{Arrow plots for our simulation and mannik's experimental}
%%   \label{arrow-compare-mannik}
%% \end{figure}

\section{Comparison of Mannik's Experimental and Our Stadium Shapes}
In general we have found that the deterministic simulations of the
protein reaction sequence discussed above results in robust
oscillatory behavoir of polar selection and oscillation in a wide
variety of cell shapes.  On the other hand, the stochastic simulations
show back and forth behavoir that breaks down in more assymetric
shapes with larger sizes, which is similar to what Mannik has observed
experimentally.

\begin{figure}
  \includegraphics[width=\columnwidth]{../data/shape-stad/plots/arrow-400-stoch}
  \caption{Arrow plots of our stadium shape}
  \label{stadium-arrow}
\end{figure}

\begin{figure}
  \includegraphics[width=\columnwidth]{../paper/plot-ave}
  %% \includegraphics[width=\columnwidth]{../data/shape-randst/plots/arrow-1500-95-stoch}
  %% \includegraphics[width=\columnwidth]{../data/shape-randst/plots/arrow-1850-95-stoch}
  %% \includegraphics[width=\columnwidth]{../data/shape-randst/plots/arrow-2500-95-stoch}
  \caption{Arrow plots of our randst shape}
  \label{stadium-arrow}
\end{figure}

%% \begin{figure}
%%   \includegraphics[width=\columnwidth]{../data/shape-randst/plots/arrow-1400-94-stoch}
%%   \includegraphics[width=\columnwidth]{../data/shape-randst/plots/arrow-1860-94-stoch}
%%   \includegraphics[width=\columnwidth]{../data/shape-randst/plots/arrow-2500-94-stoch}
%%   \caption{Arrow plots of our randst shape}
%%   \label{stadium-arrow}
%% \end{figure}

\fixme{specifically design other arrow plot (with many comparisons) to
  illustrate point}

@@

%% \section{Triangular Shapes}

%% \begin{figure}
%% %  \includegraphics[width=4.2cm]{../data/shape-triangle/plots/box-plot_D--triangle-25-233-233-233-1500}
%% %  \includegraphics[width=4.2cm]{../data/shape-triangle/plots/box-plot_D--triangle-25-266-266-266-1500}
%%   \includegraphics[width=\columnwidth]{../data/shape-triangle/plots/box-plot_D--triangle-25-233-233-233-1500}
%%   \includegraphics[width=\columnwidth]{../data/shape-triangle/plots/box-plot_D--triangle-25-266-266-266-1500}
%%   %\includegraphics[width=4.2cm]{../data/shape-triangle/plots/box-plot_D--triangle-25-800-800-800-1500}
%%   \caption{Two box plots of triangle shapes.  The two smaller show
%%     sizes at the lower stability threshhold, one just below, one just
%%     above.  The larger triangle is the largest simulated. The vertical
%%     axis shows stacked and changing over time the total number of
%%     proteins that are of four different compound stages and in the
%%     different sections of the cell.}
%%   \label{box-triangle}
%% \end{figure}

%% Along with our other shapes we have chosen to simulate both
%% equilateral and iscoscolese triangular flattened shapes, since these
%% offer in a way the simplest shapes that are geometrically more
%% complicated than our pills.

%% Both triangle types show a similar oscillation pattern.  The
%% iscosolese exhibit regular oscillations characterized by one distinct
%% pole in the more accute angle of the three, and another 'pole' along
%% the opposite wall, peaking in the two more obtuse angles at the same
%% time.  The equilateral triangles settle very quickly into the same
%% pattern, where the choice of which is one pole corner is dependent
%% upon initial conditions.  As with the Mannik simulated shapes, there
%% is a critical small size below which the oscillations dampen out and
%% will not achieve a regular pattern.  Figure~\ref{box-triangle} shows
%% two plots of equilateral triangles.  The two smaller show the
%% largest that showed damped out behavoir and the smallest that
%% developed regular oscillations.  The larger shows the largest traingle
%% that we simulated, showing that at even very large sizes the behavoir
%% is consistent and stable.  Above the small size threshold, all
%% simulated triangles show regular patterns for as long as simulated
%% (roughly 800 seconds).
